---
title: "Getting and Cleaning Data Course Project"
author: "Kelvin Leung"
date: Nov 6, 2016
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. You will be graded by your peers on a series of yes/no questions related to the project. You will be required to submit: 1) a tidy data set as described below, 2) a link to a Github repository with your script for performing the analysis, and 3) a code book that describes the variables, the data, and any transformations or work that you performed to clean up the data called CodeBook.md. You should also include a README.md in the repo with your scripts. This repo explains how all of the scripts work and how they are connected.

One of the most exciting areas in all of data science right now is wearable computing - see for example this article . Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

Here are the data for the project:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

You should create one R script called run_analysis.R that does the following.

  * Merges the training and the test sets to create one data set.

  * Extracts only the measurements on the mean and standard deviation for each measurement.

  * Uses descriptive activity names to name the activities in the data set

  * Appropriately labels the data set with descriptive variable names.

  * From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

<br>
<section>
<br>

## I. Get the data
1. Download the file and put the file in the `data` folder

```{r get data}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
destf1 <- "./data/Dataset.zip"
download.file(fileUrl, destfile = destf1, method='curl')

```

2. Unzip the file

```{r unzip}
unzip(zipfile=destf1, exdir='./data')
```

3. List out the files

```{r listfile}
path_to_file <- file.path("./data", "UCI HAR Dataset")
fs <- list.files(path_to_file, recursive = TRUE)
fs
```

<!-- ``` -->
<!-- ## Features -->

<!--  1. Triaxial acceleration from the accelerometer (total acceleration) i.e.tAcc-XYZ   and the estimated body acceleration. -->
<!--  2. Triaxial Angular velocity from the gyroscope.  i.e. tGyro-XYZ -->
<!--  3. A 561-feature vector with time and frequency domain variables.  -->
<!--  4. Its activity label.  -->
<!--  5. An identifier of the subject who carried out the experiment. -->

<!--  3 for acceleration +  -->
<!--  3 for angular velocity +  -->
<!--  561 (freq domain vars) +  -->
<!--  1 activity label +  -->
<!--  1 identifier of the subject  -->
<!--  =  total numbers of feature: 3+3+561+1+1 = 569 -->

<!--  30% for testing -->

<!--   test/subject_test.txt <- activity_lables for the testing set -->
<!--   test/X_test.txt  -->
<!--   test/y_test.txt <- activity_labels -->

<!--  70% for training -->

<!--   train/subject_train.txt <- activity_labels for the training set -->
<!--   train/X_train.txt  -->
<!--   train/y_train.txt <- activity_labels -->
<!-- ``` -->

```
The following table depicts the structure of the data frame that will be used in the project.

  ----------------  ------------------------------------------------
  variable names    | features    |     subject       | activity
  ----------------  -------------- ------------------- -------------
  
  Data              | X_train.txt | subject_train.txt | y_train.txt
                    |             |                   |
                    | X_test.txt  | subject_test.txt  | y_test.txt 
                    |             |                   |
  ------------------------------------------------------------------
  Table: Data Frame used in this project.
  
  1. Values of variable "activity" consist of data from "Y_train.txt" and "Y_test.txt"
  2. Values of variable "features" consist of data from "X_train.txt" and "X_test.txt"
  3. Values of variable "subject" consist of data from "subject_train.txt" and "subject_test.txt"
  4. Names of variable "features" come from "features.txt"
  5. Levels of variable "activity" come from "activity_labels.txt"
  
  We will use "activity", "subject", and "features" as part of descriptive variable names for the data frame. 
  
```



##II. Read Data Files

**0. A checkPercentage function** 
```{r checkPercentage, echo=TRUE}
checkPercentage <- function(test, train) {
  # a fucntion to compute the percentage of rows in test w.r.t. the sum of 
  # the row in test and tain set. return -1 if error
  # 
  # Note: use it intelligently 
  #
  ans <- -1
  if (is.data.frame(test) & is.data.frame(train)) {
    nrow.test <- nrow(test)
    nrow.train <- nrow(train)
    total_row <- nrow.test + nrow.train
    ans <- nrow.test / total_row
  }
  return (ans)
}
```

**1. Read the Activity file **

```{r read activity file}
act_train_file <- file.path(path_to_file, "train", "y_train.txt")
data_Act_train <- read.table(act_train_file, header=FALSE) 

act_test_file <- file.path(path_to_file, "test", "y_test.txt")
data_Act_test <- read.table(act_test_file, header=FALSE) # use data_Act_test
```
- 1.1 Check for "activity" data
```{r check activity data}
percent_test <- checkPercentage(data_Act_test, data_Act_train) # it is roughly 0.3
percent_train <- checkPercentage(data_Act_train, data_Act_test) # it is roughly 0.7
print.noquote(paste("For Activities: percent_train=[", round(percent_train, digits = 4), "], percent_test=[", round(percent_test, digits =4) , "], total=[", (percent_test+percent_train), "].", sep=''))
```
- 1.2 Look at the variables in data frame: **data_Act_train** (activity train set)
```{r look at the contents in data_Act_train data frame}
str(data_Act_train)
```
- 1.3 Look at the varialbles in data frame: **data_Act_test** (activity test set)
```{r look at the contents in data_Act_test data frame}
str(data_Act_test)
```

**2. Read the "subjects" file**

```{r read subjects file}
sub_train_file <- file.path(path_to_file, "train", "subject_train.txt")
data_Subject_train <- read.table(sub_train_file, header=FALSE) # use data_Subject_train

sub_test_file <- file.path(path_to_file, "test", "subject_test.txt")
data_Subject_test <- read.table(sub_test_file, header=FALSE) # use data_Subject_test
```
- 2.1 Check for "subjects" data"
```{r test_for_subject}
percent_train <- checkPercentage(data_Subject_train, data_Subject_test)
percent_test <- checkPercentage(data_Subject_test, data_Subject_train)
print.noquote(paste("For Subjects: percent_train=[", round(percent_train, digits = 4), "], percent_test=[", round(percent_test, digits =4) , "], total=[", (percent_test+percent_train), "].", sep=''))
```
- 2.2 Look at the variables in data frame: **data_Subject_train** (subjects train set)
```{r look at the contents in data_Subject_train data frame}
str(data_Subject_train)
```
- 2.3 Look at the variables in data frame: **data_Subject_test** (subjects test set)
```{r look at the contents in data_Subject_test data frame}
str(data_Subject_test)
```

**3. Read the Features files**

```{r read "features" file }
features_train_file <- file.path(path_to_file, "train", "X_train.txt")
data_Features_train <- read.table(features_train_file, header=FALSE)

features_test_file <- file.path(path_to_file, "test", "X_test.txt")
data_Features_test <- read.table(features_test_file, header=FALSE)
```
- 3.1 Check for "features" data
```{r check for "features"" data}
percent_train <- checkPercentage(data_Features_train, data_Features_test)
percent_test <- checkPercentage(data_Features_test, data_Features_train)
print.noquote(paste("For Features: percent_train=[", round(percent_train, digits = 4), "], percent_test=[", round(percent_test, digits =4) , "], total=[", (percent_test+percent_train), "].", sep=''))
```
- 3.2 Look at the variables in data frame: **data_Features_train** (features train set)
```{r look at the contents in data_Features_train data frame}
str(data_Features_train)
```
- 3.3 Look at the variables in data frame: **data_Features_test** (features test set)
```{r look at the contents in data_Features_test data frame}
str(data_Features_test)
```

##Q1: Merges the training and the test sets into `Data` data.frame

1. use rbind to concatenate rows in "activity", "subject" and "features" data frame
```{r use rbind to combine df for activity, subject and features}
data_Activity <- rbind(data_Act_train, data_Act_test)
data_Subject <- rbind(data_Subject_train, data_Subject_test)
data_Features <- rbind(data_Features_train, data_Features_test)
```

2. set names to variables 
```{r set names to variables}
names(data_Activity) <- c("activity")
names(data_Subject) <- c("subject")
```

3. read features names from features.txt file
```{r read features names}
featureNames <- read.table(file.path(path_to_file, "features.txt"), header=FALSE)$V2
names(data_Features) <- featureNames
```

4. merge columns together to get the data frame "Data" for all data
```{r merge them into a big **Data** frame}
Data <- cbind(data_Features, cbind(data_Subject, data_Activity))
```

5. check out the Data: 
```{r check out Data}
str(Data)
```

##Q2: Extracts only the measurements on the mean and standard deviation for each measurement.

1. Subset Name of Features by measurements on the mean and standard deviation.

   Use: regex for mean or standard deviation is mean\\\\(\\\\) | std\\\\(\\\\)


```{r use regex to filter out mean and standard deviation}
sub_FeaturesNames <- featureNames[grep("mean\\(\\)|std\\(\\)", featureNames)]
```

2. Subset the data.frame Data by selected names of Features
```{r subset data using selected names of features}
selected_sub_FeaturesNames_subj_act <- c(as.character(sub_FeaturesNames), "subject", "activity")
Data <- subset(Data, select=selected_sub_FeaturesNames_subj_act)
```

3. Check the data frame Data
```{r check Data}
str(Data)
```

##Q3: Uses descriptive activity names to name the activities in the data set

1. Read descriptive activity names from "activity_labels.txt"
```{r read activity_labels.txt}
labels_file <- file.path(path_to_file, "activity_labels.txt")
actLabels <- read.table(labels_file, header=FALSE)
```

2. Factorize variable "activity" in the data frame **Data** using descriptive activity names
```{r factorization}
Data$activity <- factor(Data$activity)
# use of factor with labels 
Data$activity <- factor(Data$activity, labels=as.character(actLabels$V2))
```

3. check Data$activity
```{r check Data$activity}
head(Data$activity, 30)
```

##Q4: Appropriately labels the data set with descriptive variable names.

1. Replace Features using descriptive variable names
    - replace prefix "t" by "time"
    - replace prefix "f" by "frequency"
    - replace "Acc" with "Accelerometer"
    - replace "Gyro" with "Gyroscope"
    - replace "Mag" with "Magnitude"
    - replace "BodyBody" with "Body"

```{r replace features using descriptive variable names}
names(Data) <- gsub("^t", "time", names(Data))           # replace ^t with "time"
names(Data) <- gsub("^f", "frequency", names(Data))      # replace ^f with "frequency"
names(Data) <- gsub("Acc", "Accelerometer", names(Data)) # replace Acc with Accelerometer
names(Data) <- gsub("Gyro", "Gyroscope", names(Data))    # replace Gyro with Gyroscope
names(Data) <- gsub("Mag", "Magnitude", names(Data))     # replace Mag with Magnitude
names(Data) <- gsub("BodyBody", "Body", names(Data))     # replace BodyBody with Body
```

1.1 check names(Data)
```{r check variable names of "Data" data frame}
names(Data)
```

##Q5: Creates a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r create tidy data set and output to a file}
Data2 <- aggregate(. ~subject + activity, Data, mean)
Data2 <- Data2[order(Data2$subject, Data2$activity), ]
write.table(Data2, file="tidydata.txt", row.name=FALSE)
```

```{r check Data2}
head(Data2[,1:5], n=3)
```


